Verification of concurrent design patterns with data

### 2019-04-26 2018-Arch-pNets/Coordination Camera ready/arch-pNets.tex
Almost final version

% This is LLNCS.DOC the documentation file of
% the LaTeX2e class from Springer-Verlag
% for Lecture Notes in Computer Science, version 2.4
\documentclass{llncs}
\usepackage{llncsdoc}

@@

\begin{abstract}
We provide a solution for the design of safe concurrent systems by
compositional application of verified design patterns\mdash 
called \emph{architectures}\mdash to a small set of functional
components.  To this end, we extend the theory of architectures
developed previously for the BIP framework with the elements necessary
for handling data: definition and operations on data domains, syntax
and semantics of composition operators involving data transfer.
%
\newCoord{We provide a property preservation result, which states that
%
under
monotonicity assumptions on guards and assignments and, in presence of the \emph{maximal progress} special case of
  priority models, additional upwards compatibility assumptions on the architectures,} composition of
architectures preserves their characteristic safety properties.
To verify that individual architectures do enforce their associated
properties, we provide an encoding into open pNets, an intermediate
model  that supports SMT-based verification.
%% Additionally, the translation into  open pNet  \noteEM{ICI}{gives
%% us a strong insight and a comparison of the coordination mechanisms
%% provided by the two frameworks}.
The approach is illustrated by a case
study based on a previously developed BIP model of a nanosatellite
on-board software.

@@

This section presents the extension of the theory of
  architectures \cite{AttieBBJS16-architectures-faoc} with data and
  briefly discusses a special case of priority models, called
  \emph{maximal progress}.
%
  \newCoord{These extensions require us to define the framework in a
    manner that would allow formulating and proving the property
    preservation result (\thm{combining} below).  In
    \cite{AttieBBJS16-architectures-faoc}, this result is obtained by
    requiring, in the definition of architecture composition, that an
    interaction among coordinated components be only possible if both
    architectures ``agree'' that it should be enabled.  With respect
    to data, the main difficulty lies in ensuring that this
    ``agreement'' extends to the transferred data values.  A trivial
    extension would allow an interaction only if the data values
    proposed by both architectures coincide.  Considering this
    requirement too restrictive, we go beyond by assuming the data
    domains to be ordered and taking the \emph{meet} of the proposed
    values.  The property preservation result then holds independently
    of the proposed values, provided that guards and update
    assignments are monotonic.

@@

\begin{definition}[Component semantics]
  \label{defn:comp:semantics}
  The \emph{semantics} of a component $B = \tuple{Q, q^0, V,
  \val{0}{}, P, \export, \goesto{}}$ is %% given by
  the LTS denoted $\semopen{B} = \tuple{S,
  s^0, \goesto{}}$, where $S = Q \times \valuations{V}$, $s^0 =
  \tuple{q^0, \val{0}{}}$ and $\goesto{}$ is the minimal transition
  relation satisfying the rule
  %
  \begin{equation}
    \label{eq:comp:semantics}
    \infer{
      q \goesto{a, g, u} q'
      \and
      \val{}{} \models g
      \and
      \val[\primeit]{}{} = \tilde{\val{}{}}[u]
      \and
      \valdiff{\val{}{}}{\tilde{\val{}{}}} \subseteq \export(a)
    }{
      (q, \val{}{}) \goesto{a,\tilde{\val{}{}}} (q', \val[\primeit]{}{})
    }
    \,.
  \end{equation}
  %
  %% \todoSB{Do we actually need this?}\todoLH{I agree: I do not think we need this}
  %% \hl{The \emph{closed semantics} of $B$ is given by the LTS denoted
  %% $\semclosed{B} \bydef{=}\reachable{\semopen{B}}$, comprising
  %% only the reachable states of $\semopen{B}$.}
\end{definition}

The use of the intermediate valuation $\tilde{\val{}{}}$ in the
conclusion and the third premise of  \eq{comp:semantics}
allows some  variables to get new values before the
transition is  fired.  Thus the component is \emph{open}
to the exchange of data with its environment.  However, the
fourth premise  states that only the variables exported through the ports participating in the interaction can be affected by the data transfer.
%restricts the set of
%variables that can be affected by such data transfer to those
%that are exported through the ports participating in the
%interaction.

@@

In addition to interaction models, BIP relies on 
  \emph{priority models} that impose  a strict partial order on
  interactions.  Formal definitions are provided in \app{arch:appendix}.  Intuitively, an interaction can be fired only if all
  the higher-priority interactions available in the current state are
  disabled by their respective guards.
%
In the next sections, we will implicitly assume application of the  \emph{maximal
  progress} priority $\mu$, where $(a, g, u) \prec_\mu (b, h, w)$ iff  $a \subset b$ and $a \neq b$.  For instance, the
port $\NameTimer.\PortTick$ will never fire alone if the port $\PortFail$ is also enabled.
In \secn{conclusion}, we discuss the generalisation of the
obtained results to other priority models.

@@

%\begin{proposition}[Properties of $\arcomp$]
%  \label{prop:arcomp:nice}
  %% Architecture composition
$\arcomp$ is associative and commutative.
  %%; it is idempotent if all coordinators
  %% are deterministic; $A_{id} = \bigl(\emptyset, \emptyset,
  %% \emptyset, \emptyset, \{(\emptyset, \true, \noop)\}\bigr)$ is its neutral element, \ie for
  %% any architecture $A$, we have $A \oplus A_{id} \arequiv A$.
  %% Furthermore, for any component $B$, we have $A_{id}(B) = \mu(B)$.
%\end{proposition}
%
%% \begin{proof}[Sketch of the proof]
%%   %% Commutativity and associativity 
%%   Follows from the corresponding
%%   properties of set union, Boolean conjunction and
%%   semilattice meet.
%% %
%% %% \todoLH{Check when equivalence is defined}
%% %%   Suppose we have two architectures $A \arequiv A'$.  This does not
%% %%   necessarily mean that their sets of coordinators
%% %%   coincide.  However, if all the involved coordinators
%% %%   are deterministic, then, in any state of $(A \arcomp A')(\cB)$,
%% %%   both architectures will impose the same restrictions, enabling
%% %%   the same interactions between the coordinators and operand
%% %%   components.  Hence, we have $(A \arcomp A')(\cB) = A(\cB) =
%% %%   A'(\cB)$.  Since this holds for any set of components $\cB$, we
%% %%   conclude that $A \arcomp A' \arequiv A \arequiv A'$.
%% %% %
%% %%   The properties of $A_{id}$ follow immediately from the
%% %%   definitions of architecture application and composition.
%% \end{proof}
%% The last statement of \prop{arcomp:nice} highlights a subtle point:
%% $A_{id}$ is the neutral element of the $\arcomp$ operator\mdash it is
%% not the identity operator on components.
%****************************************************************
%\subsubsection{Preservation of safety properties}
%\label{secn:safety}
For a component $B$, we denote $S_{\semopen{B}}$ and
$s^0_{\semopen{B}}$ the corresponding constituents of
$\semopen{B}$ (see Definition~\ref{defn:comp:semantics}).

@@

\begin{definition}[Properties]
  \label{defn:property}
  Let $B$ be a component.  A \emph{(safety) property} 
  of $B$ is a predicate $\Phi$ on
  $S_{\semopen{B}}$, such that $\bigl((q,
  \val{}{}) \models \Phi\bigr) \land (\val[\primeit]{}{} \order
  \val{}{})$ implies $(q, \val[\primeit]{}{}) \models \Phi$.
  %% ,
  %% where we write $(q, \val{}{}) \models \Phi$ iff $\Phi(q,
  %% \val{}{}) = \true$.
  A property $\Phi$ is \emph{initial} if
  $s^0_{\semopen{B}} \models \Phi$.
  %% \todoSB{Do we need this?}\hl{;
  %% it is \emph{reachable} iff there exists a possibly empty path
  %% $s^0_{\semopen{B}} \goesto{a^1, \val{1}{}} s^1 \goesto{a^2,
  %%   \val{2}{}} \cdots \goesto{a^n, \val{n}{}} s^n$, such that
  %% $s^n \models \Phi$}.
\end{definition}

\label{property:reset}
Although we define properties as state predicates, any
appropriate logic can be used to specify them.  For instance, the
property \emph{``There is always a possibility to reset the system
  after a single failure''} (\ie without
  additional failures having to occur in the meantime) enforced by the
  Failure Monitor architecture \newCoord{comprises the safety component that} can be specified using CTL as
  %  \noteSB{Removed $\lnot \PortResume$, because not needed \emph{and} could not reformulate nicely in English.}
    $\AG
\bigl(\PortFail \rightarrow \EX \EW[]{\lnot \PortFail}{
  \PortReset}\bigr)$.  Additional properties are provided in
\app{properties}.
%% \todoSB{Cite the original paper somewhere.}
An architecture enforces
its characteristic property on  its operand components.
From this point of view, the set of coordinators is
not relevant, neither are their states.  Thus, 
properties enforced by architectures only involve
the unrestricted composition of the operands:

\begin{definition}[Enforcing properties]
  \label{defn:enforce}
  Let $A = \tuple{\cC, P_A, V_A, \export[A], \Gamma}$ be an architecture; let $\cB$
  be a set of components and $\Phi$ an initial property of
  their parallel composition $\Gamma_{\|}(\cB)$,
  with $\Gamma_{\|} = \setdef{(a,\true,\emptyset)}{a \subseteq \bigcup_{B \in \cB} P_B}$.
  We say that \emph{$A$ enforces $\Phi$ on
    $\cB$} iff, for every state $s = (s_c, s_b)$ reachable in
  $\semopen{A(\cB)}$, with 
  $s_c \in \prod_{C \in \cC} S_{\semopen{C}}$ and
  $s_b \in \prod_{B \in \cB} S_{\semopen{B}}$,
  we have $s_b \models \Phi$.
\end{definition}
In the following, when we say that an
architecture enforces some property $\Phi$, $\Phi$ is supposed to be initial for the coordinated components.
%
%According to the above definition, when we say that an
%architecture enforces some property $\Phi$, it is implicitly
%assumed that $\Phi$ is initial for the coordinated components.
%Below, we omit mentioning this explicitly.
\defn{up:compat} in \app{arch:appendix} formally defines
 \emph{upwards compatibility} that ensures property preservation when composing architectures. Informally, two architectures $A_1$ and $A_2$
are  upwards compatible iff, whenever their composition
involves the fusion of two interactions $a_1 = a \cap P_{A_1}$ and
$a_2 = a \cap P_{A_2}$ (see \eq{arch:composition}) and one, say
$a_1$, is inhibited in a given state by a larger interaction $b_1
\supset a_1$, there exists an interaction $b_2 \supseteq a_2$ that can
be fused with $b_1$ to form an interaction enabled in the same state.

@@

  Although somewhat technical, the encoding of coordinators is, in
  fact, pretty straightforward, comprising three key ideas: 1)~we
  introduce an additional transition (hence also an additional state)
  to explicitly initialise the variables; 2)~we introduce additional
  input variables to manipulate the values provided to the coordinator
  by the rest of the system for all exported variables; and
  3)~following the classical technique~\cite{milner83-calculi}, we
  simulating absence of action by an additional loop transition.

  The encoding of connectors (interaction models) is more involved.
  Since a connector represents a set of potential interactions, some
  ports may not participate in all of them.  To encode this
  possibility, we introduce, for each port, an additional Boolean
  variable denoting whether the port participates in the interaction
  or not and, for each connector, a predicate characterising the
  interaction pattern.  Intuitively, the semantics of \emph{flat} BIP
  connectors~\cite{BliSif08-acp-tc} depends on the synchron/trigger
  annotations of ports.  If all ports in a connector are synchrons,
  the only allowed interaction is that comprising all the ports, \ie
  they all have to ``agree to interact''.  This case corresponds
  precisely to the semantics of synchronisation vectors in pNets.  If
  a connector has at least one trigger, then the allowed interactions
  are those that comprise at least one trigger, \ie they must be
  ``initiated by a trigger''.  In a hierarchical connector, these
  principles are applied recursively.  Thus, we observe a
  ``causality'' relation among ports of a connector: participation of
  a synchron in an interaction \emph{implies} that of a trigger.
  Causal Interaction Trees and Systems of Causal Rules, proposed in
  \cite{BliSif10-causal-fmsd}, formalise this causality relation and
  provide transformations from connectors to Boolean predicates and
  back.  Since we use SMT techniques for the analysis of the resulting
  pNet, the encoding presented below is optimised to reduce the number
  of variables by treating separately the ``top-level'' triggers, \ie
  those thereof the participation in an interaction is not subject to
  that of any other ports.
%
}

@@

In the general case, the encoding relies on the causal semantics of the
algebra of BIP connectors \cite{BliSif10-causal-fmsd}.  Disregarding
the variables and data transfer, the Algebra of Connectors $\ac(P)$
\cite{BliSif07-acp-emsoft} provides a syntactic notation for the BIP
connectors.  The causal semantics of the connectors, given in terms of
the Algebra of Causal Interaction Trees $\ct(P)$, explicitates the
causal dependencies through an encoding mapping
$\tau: \ac(P) \rightarrow \ct(P)$.  Another mapping $R: \ct(P)
\rightarrow \cru(P)$ encodes causal interaction trees into systems of
causal rules, which are Boolean implications similar to the ones in
the example above.  We briefly summarise the key definitions and
results in \app{algebras}.  The $\ac(P)$, $\ct(P)$ and $\cru(P)$
representations of the connectors in \fig{schema:ArchFailure:BIP} are
shown in \tab{connectors} (elements shown in red can be removed for
simplification as described in the example above).

\begin{table}[t]
  \caption{Algebraic representations of the connectors in \fig{schema:ArchFailure:BIP} (see \app{algebras})}
  \label{tab:connectors}
  \resizebox{\columnwidth}{!}{%
  \begin{tabular}{@{}c|c|c@{}}
    \hline {\bf Connector} & {\bf Causal Interaction Tree} & {\bf
      System of Causal Rules}
    \\\hline    
%
    \begin{tikzpicture}[shorten >=1pt,node distance=.7cm,>=stealth']

@@

Now we lift this encoding to the data-sensitive case. Below, we assume that, as in \fig{schema:ArchFailure:BIP}, the
interaction model is defined by a set of connectors, annotated with
Boolean guards and with update assignments.  In particular, we assume
that the guards and update assignments are well-defined for any
interaction allowed by the connector.  For example, the choice
$\NameCtrl.\PortFail\ ?\ \NameCtrl.\mathit{zone} : \topinterval$ in
the update assignment
$
\NameTimer.\mathit{z} := \NameTimer.\mathit{z}
\cap \bigl(
(
\NameCtrl.\PortFail\ ?\ \NameCtrl.\mathit{zone} : \topinterval
)
+ \NameTimer.\mathit{t} \bigr)
$
associated to the connector {\NameTimer.\PortTick \trigsynch
  (\PortFail \trigsynch \NameCtrl.\PortFail)} in
\fig{schema:ArchFailure:BIP} ensures that the assignment is
well-defined independently of whether {\NameCtrl.\PortFail}
participates or not.  Let us denote by $\gamma \subset \ac(P_A)
\times \guards{V_A} \times \assigns{V_A}$ the set of connectors in the
architecture $A$ and by $P_x$ the set of ports involved in the
connector $x \in \gamma$.  Then,
%\addSB{
the interaction model defined by $\gamma$ is 
  $\Gamma = \setdef{(a,g,u)}{(x,g,u) \in \gamma, a \in \intsem{x}}$
and the set of synchronisation vectors encoding $\Gamma$ is
%}
$\nopri[\Gamma] \bydef{=}
\setdef{\nopri[x,g,u]}{(x,g,u) \in \gamma}$, with %% $\nopri[x,g,u]$ is the  defined by
%
\begin{multline}
  \label{eq:nopri:connector}
  \nopri[x,g,u] \bydef{=}
  %% \Tuple{p\bigl(\export[A](p),{{\color{blue}\export[A]'(p),}}b_p\bigr)}^{p \in P_x}
  \Tuple{\Setdef{p\bigl(\export[A](p),\export[A]'(p),b_p\bigr)}{p \in P_x \cap P}}^{P \in P_C^{C \in \cC} \cup \partition}
  \rightarrow
  \alpha \bigl(b_p^{p \in P_x}\bigr)
  \\
  \left[
    \bigwedge R(\tau(x))\bigl[\nicefrac{b_p}{p}\bigr]
    \land
    \bigwedge_{(x := e_x) \in u} (?x' = e_x)
    \land
    \bigwedge_{x \in \export[A](x), x \not\in u} (?x' = x)
    \right]
  \,,
\end{multline}
%
with $b_p^{p \in P_x}$ fresh Boolean variables, $\alpha$ a fresh name, $\tau :
\ac(P_A) \rightarrow \ct(P_A)$ and $R : \ct(P_A) \rightarrow
\cru(P_A)$ the two mappings defined in \app{transformations} and
illustrated in \tab{connectors}, $\bigl[\nicefrac{b_p}{p}\bigr]$ is
the substitution that replaces in the expression that precedes it
all occurrences of all $p$ by corresponding variables $b_p$.

%\addSB{%
%
For the sake of clarity, we simplify the case study encoding in
\fig{FailureTimer:pNet} by reusing the port names of the original
architecture instead of fresh names $\alpha$.  This is made
possible by the fact that each synchronisation vector involves at
most one action of interest (see the properties in
\app{properties}).
%% In the general encoding, the global actions of the
%% synchronisation vectors should be fresh action names.
%% However,
%% for the properties considered in our case study, we only need to
%% exhibit the actions of the holes.  Furthermore, each
%% synchronisation vector involves at most one such action.
%% This
%% allows us to reuse them as the global actions of the
%% synchronisation vectors, simplifying the translation of
%% properties to be verified (\cf \app{properties}).
%
%}\noteLH{ce
  %% paragraphe sonne vraiment note de rapport de recherche, tu
  %% arrives a lui donner un peu d'abstraction/de recul?}\noteSB{Est-ce que c'est mieux comme \c{c}a ? OUI -- Ludo}

>>>>

% This is LLNCS.DOC the documentation file of
% the LaTeX2e class from Springer-Verlag
% for Lecture Notes in Computer Science, version 2.4
\documentclass{llncs}
\usepackage{llncsdoc}
\usepackage[T1]{fontenc}

@@

\begin{abstract}
We provide a solution for the design of safe concurrent systems by
compositional application of verified design patterns\mdash 
called \emph{architectures}\mdash to a small set of functional
components.  To this end, we extend the theory of architectures
developed previously for the BIP framework with the elements necessary
for handling data: definition and operations on data domains, syntax
and semantics of composition operators involving data transfer.
%
\newCoord{We provide a set of conditions under which} %% property preservation result holds that states that
%% %
%% under
%% monotonicity assumptions on guards and assignments and, in presence of the \emph{maximal progress} special case of
%%   priority models, additional upwards compatibility assumptions on the architectures,}
composition of
architectures preserves their characteristic safety properties.
To verify that individual architectures do enforce their associated
properties, we provide an encoding into open pNets, an intermediate
model  that supports SMT-based verification.
%% Additionally, the translation into  open pNet  \noteEM{ICI}{gives
%% us a strong insight and a comparison of the coordination mechanisms
%% provided by the two frameworks}.
The approach is illustrated by a case
study based on a previously developed BIP model of a nanosatellite
on-board software.

@@

This section presents the extension of the theory of
  architectures \cite{AttieBBJS16-architectures-faoc} with data and
  briefly discusses a special case of priority models, called
  \emph{maximal progress}.
%
  \newCoord{These extensions require us to define the framework in a
    manner that would allow formulating and proving the property
    preservation result (\thm{combining} below).  In
    \cite{AttieBBJS16-architectures-faoc}, this result is obtained by
    requiring, in the definition of architecture composition, that an
    interaction among coordinated components be only possible if both
    architectures ``agree'' that it should be enabled.  With respect
    to data, the main difficulty lies in ensuring that this
    ``agreement'' extends to the transferred data values.  A trivial
    extension would allow an interaction only if the data values
    proposed by both architectures coincide.  As this
    requirement is too restrictive, we go beyond by assuming the data
    domains to be ordered and taking the \emph{meet} of the proposed
    values.  The property preservation result then holds independently
    of the proposed values, provided that guards and update
    assignments are monotonic.

@@

\begin{definition}[Component semantics]
  \label{defn:comp:semantics}
  The \emph{open semantics} of a component $B = \tuple{Q, q^0, V,
  \val{0}{}, P, \export, \goesto{}}$ is %% given by
  the LTS denoted $\semopen{B} = \tuple{S,
  s^0, \goesto{}}$, where $S = Q \times \valuations{V}$, $s^0 =
  \tuple{q^0, \val{0}{}}$ and $\goesto{}$ is the minimal transition
  relation satisfying the rule
  %
  \begin{equation}
    \label{eq:comp:semantics}
    \infer{
      q \goesto{a, g, u} q'
      \and
      \val{}{} \models g
      \and
      \val[\primeit]{}{} = \tilde{\val{}{}}[u]
      \and
      \valdiff{\val{}{}}{\tilde{\val{}{}}} \subseteq \export(a)
    }{
      (q, \val{}{}) \goesto{a,\tilde{\val{}{}}} (q', \val[\primeit]{}{})
    }
    \,.
  \end{equation}
%
  The \emph{closed semantics} of $B$ is given by the LTS denoted
  $\semclosed{B}$, comprising only those transitions of $\semopen{B}$,
  where $\tilde{\val{}{}} = \val{}{}$, which we denote
  $(q, \val{}{}) \goesto{a} (q', \val[\primeit]{}{})$.
\end{definition}

The use of the intermediate valuation $\tilde{\val{}{}}$ in the
conclusion and the third premise of  \eq{comp:semantics}
allows some  variables to get new values before the
transition is  fired.  Thus the component is \emph{open}
to the exchange of data with its environment.  However, the
fourth premise  states that only the variables exported through the ports participating in the interaction can be affected by the data transfer.
%restricts the set of
%variables that can be affected by such data transfer to those
%that are exported through the ports participating in the
%interaction.
%
The closed semantics excludes this possibility of data exchange.

@@

In addition to interaction models, BIP relies on 
  \emph{priority models} that impose  a strict partial order on
  interactions.  %% Formal definitions are provided in \app{arch:appendix}.
  Intuitively, an interaction can be fired only if all
  the higher-priority interactions available in the current state are
  disabled by their respective guards.
%
In the next sections, we will implicitly assume application of the  \emph{maximal
  progress} priority $\mu$, where $(a, g, u) \prec_\mu (b, h, w)$ iff  $a \subset b$ and $a \neq b$.  For instance, the
port $\NameTimer.\PortTick$ will never fire alone if the port $\PortFail$ is also enabled.
%% In \secn{conclusion}, we discuss the generalisation of the
%% obtained results to other priority models.

@@

%\begin{proposition}[Properties of $\arcomp$]
%  \label{prop:arcomp:nice}
  %% Architecture composition
$\arcomp$ is associative and commutative.
  %%; it is idempotent if all coordinators
  %% are deterministic; $A_{id} = \bigl(\emptyset, \emptyset,
  %% \emptyset, \emptyset, \{(\emptyset, \true, \noop)\}\bigr)$ is its neutral element, \ie for
  %% any architecture $A$, we have $A \oplus A_{id} \arequiv A$.
  %% Furthermore, for any component $B$, we have $A_{id}(B) = \mu(B)$.
%\end{proposition}
%
%% \begin{proof}[Sketch of the proof]
%%   %% Commutativity and associativity 
%%   Follows from the corresponding
%%   properties of set union, Boolean conjunction and
%%   semilattice meet.
%% %
%% %% \todoLH{Check when equivalence is defined}
%% %%   Suppose we have two architectures $A \arequiv A'$.  This does not
%% %%   necessarily mean that their sets of coordinators
%% %%   coincide.  However, if all the involved coordinators
%% %%   are deterministic, then, in any state of $(A \arcomp A')(\cB)$,
%% %%   both architectures will impose the same restrictions, enabling
%% %%   the same interactions between the coordinators and operand
%% %%   components.  Hence, we have $(A \arcomp A')(\cB) = A(\cB) =
%% %%   A'(\cB)$.  Since this holds for any set of components $\cB$, we
%% %%   conclude that $A \arcomp A' \arequiv A \arequiv A'$.
%% %% %
%% %%   The properties of $A_{id}$ follow immediately from the
%% %%   definitions of architecture application and composition.
%% \end{proof}
%% The last statement of \prop{arcomp:nice} highlights a subtle point:
%% $A_{id}$ is the neutral element of the $\arcomp$ operator\mdash it is
%% not the identity operator on components.
%****************************************************************
%\subsubsection{Preservation of safety properties}
%\label{secn:safety}

@@

For a component $B$, we denote $S_{\semclosed{B}}$ and
$s^0_{\semclosed{B}}$ the corresponding constituents of
$\semclosed{B}$ (see Definition~\ref{defn:comp:semantics}).

\begin{definition}[Properties]
  \label{defn:property}
  Let $B$ be a component.  A \emph{(safety) property} 
  of $B$ is a predicate $\Phi$ on
  $S_{\semclosed{B}}$, such that $\bigl((q,
  \val{}{}) \models \Phi\bigr) \land (\val[\primeit]{}{} \order
  \val{}{})$ implies $(q, \val[\primeit]{}{}) \models \Phi$.
  %% ,
  %% where we write $(q, \val{}{}) \models \Phi$ iff $\Phi(q,
  %% \val{}{}) = \true$.
  A property $\Phi$ is \emph{initial} if
  $s^0_{\semclosed{B}} \models \Phi$.
  %% \todoSB{Do we need this?}\hl{;
  %% it is \emph{reachable} iff there exists a possibly empty path
  %% $s^0_{\semclosed{B}} \goesto{a^1, \val{1}{}} s^1 \goesto{a^2,
  %%   \val{2}{}} \cdots \goesto{a^n, \val{n}{}} s^n$, such that
  %% $s^n \models \Phi$}.
\end{definition}

\label{property:reset}
Although we define properties as state predicates, any
appropriate logic can be used to specify them.  For instance, the
property \emph{``There is always a possibility to reset the system
  after a single failure''} (\ie without
  additional failures having to occur in the meantime) enforced by the
  Failure Monitor architecture \newCoord{comprises the safety component that} can be specified using CTL as
  %  \noteSB{Removed $\lnot \PortResume$, because not needed \emph{and} could not reformulate nicely in English.}
    $\AG
\bigl(\PortFail \rightarrow \EX \EW[]{\lnot \PortFail}{
  \PortReset}\bigr)$.  %% Additional properties are provided in
%% \app{properties}.
%% \todoSB{Cite the original paper somewhere.}
An architecture enforces
its characteristic property on  its operand components.
From this point of view, the set of coordinators is
not relevant, neither are their states.  Thus, 
properties enforced by architectures only involve
the unrestricted composition of the operands:

\begin{definition}[Enforcing properties]
  \label{defn:enforce}
  Let $A = \tuple{\cC, P_A, V_A, \export[A], \Gamma}$ be an architecture; let $\cB$
  be a set of components and $\Phi$ an initial property of
  their parallel composition $\Gamma_{\|}(\cB)$,
  with $\Gamma_{\|} = \setdef{(a,\true,\emptyset)}{a \subseteq \bigcup_{B \in \cB} P_B}$.
  We say that \emph{$A$ enforces $\Phi$ on
    $\cB$} iff, for every state $s = (s_c, s_b)$ reachable in
  $\semclosed{A(\cB)}$, with 
  $s_c \in \prod_{C \in \cC} S_{\semclosed{C}}$ and
  $s_b \in \prod_{B \in \cB} S_{\semclosed{B}}$,
  we have $s_b \models \Phi$.
\end{definition}
In the following, when we say that an
architecture enforces some property $\Phi$, $\Phi$ is supposed to be initial for the coordinated components.
%
%According to the above definition, when we say that an
%architecture enforces some property $\Phi$, it is implicitly
%assumed that $\Phi$ is initial for the coordinated components.
%Below, we omit mentioning this explicitly.
In \cite{arch-pNets-RR}, we formally define
 \emph{upwards compatibility} that ensures property preservation when composing architectures. Informally, two architectures $A_1$ and $A_2$
are  upwards compatible iff, whenever their composition
involves the fusion of two interactions $a_1 = a \cap P_{A_1}$ and
$a_2 = a \cap P_{A_2}$ (see \eq{arch:composition}) and one, say
$a_1$, is inhibited in a given state by a larger interaction $b_1
\supset a_1$, there exists an interaction $b_2 \supseteq a_2$ that can
be fused with $b_1$ to form an interaction enabled in the same state.

@@

  Although somewhat technical, the encoding of coordinators is, in
  fact, pretty straightforward, comprising three key ideas: 1)~we
  introduce an additional transition (hence also an additional state)
  to explicitly initialise the variables; 2)~we introduce additional
  input variables to manipulate the values provided to the coordinator
  by the rest of the system for all exported variables; and
  3)~following the classical technique~\cite{milner83-calculi}, we
  simulate the absence of action by an additional loop transition.

  The encoding of connectors (interaction models) is more involved.
  Since a connector represents a set of potential interactions, some
  ports may not participate in all of them.  To encode this
  possibility, we introduce, for each port, an additional Boolean
  variable denoting whether the port participates in the interaction
  or not and, for each connector, a predicate characterising the
  interaction pattern.  Intuitively, the semantics of \emph{flat} BIP
  connectors~\cite{BliSif08-acp-tc} depends on the synchron/trigger
  annotations of ports.  If all ports in a connector are synchrons,
  the only allowed interaction is that comprising all the ports, \ie
  they all have to ``agree to interact''.  This case corresponds
  precisely to the semantics of synchronisation vectors in pNets.  If
  a connector has at least one trigger, then the allowed interactions
  are those that comprise at least one trigger, \ie they must be
  ``initiated by a trigger''.  In a hierarchical connector, these
  principles are applied recursively.  Thus, we observe a
  ``causality'' relation among ports of a connector: participation of
  a synchron in an interaction \emph{implies} that of a trigger.
  Causal Interaction Trees and Systems of Causal Rules, proposed in
  \cite{BliSif10-causal-fmsd}, formalise this causality relation and
  provide transformations from connectors to Boolean predicates and
  back.  Since we use SMT techniques for the analysis of the resulting
  pNet, the encoding presented below is optimised to reduce the number
  of variables by treating separately the ``top-level'' triggers (\eg
  $\NameTimer.\PortTick$ in the connector {\NameTimer.\PortTick
    \trigsynch (\PortFail \trigsynch \NameCtrl.\PortFail)} in
  \fig{schema:ArchFailure:BIP}).
%
}

@@

In the general case, the encoding relies on the causal semantics of the
algebra of BIP connectors \cite{BliSif10-causal-fmsd}.  Disregarding
the variables and data transfer, the Algebra of Connectors $\ac(P)$
\cite{BliSif07-acp-emsoft} provides a syntactic notation for the BIP
connectors.  The causal semantics of the connectors, given in terms of
the Algebra of Causal Interaction Trees $\ct(P)$, explicitates the
causal dependencies through an encoding mapping
$\tau: \ac(P) \rightarrow \ct(P)$.  Another mapping $R: \ct(P)
\rightarrow \cru(P)$ encodes causal interaction trees into systems of
causal rules, which are Boolean implications similar to the ones in
the example above.  %% We briefly summarise the key definitions and
%% results in \app{algebras}.
The $\ac(P)$, $\ct(P)$ and $\cru(P)$
representations of the connectors in \fig{schema:ArchFailure:BIP} are
shown in \tab{connectors} (elements shown in red can be removed for
simplification as described in the example above).

\begin{table}[t]
  \caption{Algebraic representations of the connectors in \fig{schema:ArchFailure:BIP}}
  \label{tab:connectors}
  \resizebox{\columnwidth}{!}{%
  \begin{tabular}{@{}c|c|c@{}}
    \hline {\bf Connector} & {\bf Causal Interaction Tree} & {\bf
      System of Causal Rules}
    \\\hline    
%
    \begin{tikzpicture}[shorten >=1pt,node distance=.7cm,>=stealth']

@@

Now we lift this encoding to the data-sensitive case. Below, we assume that, as in \fig{schema:ArchFailure:BIP}, the
interaction model is defined by a set of connectors, annotated with
Boolean guards and with update assignments.  In particular, we assume
that the guards and update assignments are well-defined for any
interaction allowed by the connector.  For example, the choice
$\NameCtrl.\PortFail\ ?\ \NameCtrl.\mathit{zone} : \topinterval$ in
the update assignment
$
\NameTimer.\mathit{z} := \NameTimer.\mathit{z}
\cap \bigl(
(
\NameCtrl.\PortFail\ ?\ \NameCtrl.\mathit{zone} : \topinterval
)
+ \NameTimer.\mathit{t} \bigr)
$
associated to the connector {\NameTimer.\PortTick \trigsynch
  (\PortFail \trigsynch \NameCtrl.\PortFail)} in
\fig{schema:ArchFailure:BIP} ensures that the assignment is
well-defined independently of whether {\NameCtrl.\PortFail}
participates or not.  Let us denote by $\gamma \subset \ac(P_A)
\times \guards{V_A} \times \assigns{V_A}$ the set of connectors in the
architecture $A$ and by $P_x$ the set of ports involved in the
connector $x \in \gamma$.  Then,
%\addSB{
the interaction model defined by $\gamma$ is 
  $\Gamma = \setdef{(a,g,u)}{(x,g,u) \in \gamma, a \in \intsem{x}}$
and the set of synchronisation vectors encoding $\Gamma$ is
%}
$\nopri[\Gamma] \bydef{=}
\setdef{\nopri[x,g,u]}{(x,g,u) \in \gamma}$, with %% $\nopri[x,g,u]$ is the  defined by
%
\begin{multline}
  \label{eq:nopri:connector}
  \nopri[x,g,u] \bydef{=}
  %% \Tuple{p\bigl(\export[A](p),{{\color{blue}\export[A]'(p),}}b_p\bigr)}^{p \in P_x}
  \Tuple{\Setdef{p\bigl(\export[A](p),\export[A]'(p),b_p\bigr)}{p \in P_x \cap P}}^{P \in P_C^{C \in \cC} \cup \partition}
  \rightarrow
  \alpha \bigl(b_p^{p \in P_x}\bigr)
  \\
  \left[
    \bigwedge R(\tau(x))\bigl[\nicefrac{b_p}{p}\bigr]
    \land
    \bigwedge_{(x := e_x) \in u} (?x' = e_x)
    \land
    \bigwedge_{x \in \export[A](x), x \not\in u} (?x' = x)
    \right]
  \,,
\end{multline}
%
with $b_p^{p \in P_x}$ fresh Boolean variables, $\alpha$ a fresh name, $\tau :
\ac(P_A) \rightarrow \ct(P_A)$ and $R : \ct(P_A) \rightarrow
\cru(P_A)$ the two mappings \cite{BliSif10-causal-fmsd} discussed above and
illustrated in \tab{connectors}, $\bigl[\nicefrac{b_p}{p}\bigr]$ is
the substitution that replaces in the expression that precedes it
all occurrences of all $p$ by corresponding variables $b_p$.

%\addSB{%
%
For the sake of clarity, we simplify the case study encoding in
\fig{FailureTimer:pNet} by reusing the port names of the original
architecture instead of fresh names $\alpha$.  This is made
possible by the fact that each synchronisation vector involves at
most one action of interest (see the properties in
\cite{arch-pNets-RR}).
%% In the general encoding, the global actions of the
%% synchronisation vectors should be fresh action names.
%% However,
%% for the properties considered in our case study, we only need to
%% exhibit the actions of the holes.  Furthermore, each
%% synchronisation vector involves at most one such action.
%% This
%% allows us to reuse them as the global actions of the
%% synchronisation vectors, simplifying the translation of
%% properties to be verified (\cf \app{properties}).
%
%}\noteLH{ce
  %% paragraphe sonne vraiment note de rapport de recherche, tu
  %% arrives a lui donner un peu d'abstraction/de recul?}\noteSB{Est-ce que c'est mieux comme \c{c}a ? OUI -- Ludo}

\newCoord{In the following theorem, we claim an \emph{isomorphism}
  between the open automaton semantics of a pNet encoding a BIP
  architecture and the LTS semantics of this architecture applied to a
  set of simple components.  We omit the formal definition of this
  isomorphism relation.  However, noting that open automata are,
  essentially, symbolic representations of automata with data, we can
  summarise it as follows: a transition belongs to the LTS iff a
  corresponding open transition belongs to the open automaton and the
  source and target data values of the LTS transition satisfy the
  predicate and implement the assignments of the open transition.
%
}
